\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-12-05 chapters/bci.tex
Add odd option to full page fig

\fullpagefig[caption.north]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  \caption[The \acs{bci} loop.]{%
    The \ac{bci} loop.
    The user interacts with the \ac{bci} via a specific paradigm, in this case involving
    visual stimulation.
    Electrical neural signals are recorded, and neurophysiological features related to the
    paradigm are extracted.
    Using machine learning, a decision can then be made based on these features, which can
    be presented back to the user.
    In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
  }
  \label{fig:bci/loop}
}

@@

\fullpagefig[caption.north]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and spatial resolution.
  }
  \label{fig:bci/recording}
}

>>>>

\fullpagefig[caption.north,odd]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  \caption[The \acs{bci} loop.]{%
    The \ac{bci} loop.
    The user interacts with the \ac{bci} via a specific paradigm, in this case involving
    visual stimulation.
    Electrical neural signals are recorded, and neurophysiological features related to the
    paradigm are extracted.
    Using machine learning, a decision can then be made based on these features, which can
    be presented back to the user.
    In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
  }
  \label{fig:bci/loop}
}

@@

\fullpagefig[caption.north, odd]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of \\ the skull.
	  There is a trade-off between invasiveness \\ and spatial resolution.
  }
  \label{fig:bci/recording}
}

### 2024-12-05 chapters/wcble.tex
Add odd option to full page fig

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    The \acf{wcble} training phase.
    (1) The first-stage classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs, yielding scores
    and class probabilities over time.
    (3) The medians of the probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned by shifting in time based on their corresponding class latencies by
    shifting in time.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) On convergence, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    The \acf{wcble} test phase. (1) The
	  first-stage classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs, yielding scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}

@@

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig[anchor.north]{%
  \hspace{1cm}\inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  \caption[Simulated target and non-target evoked data.]{%
    Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
    and their average (red: target, yellow: non-target, shaded area: standard
    deviation).
    Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
    spatiotemporal noise were simulated using a forward head model, and channel C3 is
    plotted here.
  }
  \label{fig:wcble/sim-sine}
}

>>>>

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig[caption.fullwidth, odd]{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    The \acf{wcble} training phase.
    (1) The first-stage classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs, yielding scores
    and class probabilities over time.
    (3) The medians of the probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned by shifting in time based on their corresponding class latencies by
    shifting in time.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) On convergence, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    The \acf{wcble} test phase. (1) The
	  first-stage classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs, yielding scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}

@@

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig[anchor.north]{%
  \hspace{1cm}\inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  \caption[Simulated target and non-target evoked data.]{%
    Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
    and their average (red: target, yellow: non-target, shaded area: standard
    deviation).
    Targets contain a sine wave pulse, jittered by $\sigma$. 32 \ac{eeg}
    channels with varying levels of noise were simulated using a forward model, channel C3 is
    plotted here.
  }
  \label{fig:wcble/sim-sine}
}
